Equation-free analysis of a dynamically evolving multigraph

### 2016-04-28 manuscript-materials/tex/response.tex

\vspace{3mm}

\textbf{First Editor's recommendations}
\vspace*{\baselineskip}

* {\em ... I would like to make a few suggestions to the authors of
  the paper for a possible extension of their methodology in the
  investigation of more realistic models of complex networks and
  possibly models supported by concrete empirical data. The notion of
  a multigraph is too general to include the whole range of
  applications of graph theory in complex network science. As a matter
  of fact, a further step would be towards the consideration of
  attributes (or labels) on nodes and links of a (multi)graph... }.

\textbf{OUR REPLY}: It is a very good suggestion to apply the method
to other systems with labeled nodes, and we have in fact already
addressed a first example of this in \cite{kattis_modeling_2016} in which
an SIS epidemiological model is investigated with similar
techniques. We have added a reference to this paper in the text.
\vspace*{\baselineskip}

### 2016-04-28 manuscript-materials/tex/multigraph-manuscript.tex
Removed comments from manuscript, fixed akis citation

\documentclass[epjST, final]{svjour} \usepackage{graphicx, subcaption,
  setspace, indentfirst, amsfonts, amsthm, empheq, color, soul}
% \usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
% \usepackage[center]{titlesec}
\graphicspath{ {../figs/} }
% \pagestyle{plain} \renewcommand\thesection{\Roman{section}}
% \renewcommand\thesubsection{\alph{subsection}.}
\newcommand\ignore[1]{}
% \setlength{\parskip}{0ex}
% \def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
% \let\endchangemargin=\endlist

\title{Equation-free analysis of a dynamically evolving multigraph}
\author{Alexander Holiday \inst{1} \and Ioannis G. Kevrekidis
  \inst{1,2}} \institute{Department of Chemical and Biological
  Engineering, Princeton University, Princeton, New Jersey 08544, USA
  \and PACM, Princeton University, Princeton, New Jersey 08544, USA}
% $^1$\textit{Department of Chemical and Biological Engineering},
% \textit{Princeton University, Princeton, New Jersey 08544, USA} \\ \\
% $^2$\textit{Department of Mathematics}, \\
% \textit{The University of British Columbia, Vancouver, British
% Columbia V6T 1Z4} } \date{}

@@

  We study the edge-conservative preferential attachment model, a
  detailed mathematical analysis of which can be found in
  \cite{rath_time_2012} and \cite{rath_multigraph_2012}.
%
  The system evolves in discrete steps $t = 0,1,\ldots t_f$, and we
  denote the $n$-vertex graph at each point by $G_n(t)$.
%
  The initial state, $G_n(0)$, is an arbitrary distribution of $m$
  edges among the $n$ vertices; the total number of both edges and
  vertices is conserved.
%
  No restrictions are placed on this initial distribution: multiple
  edges and loops are permitted. 
  % \ignore{an Erd\H{o}s-R\'{e}nyi random graph on $n$ vertices and
  % $m$ edges. Here, however, $m$ may be larger than the total number
  % of edges in a completely-connected $n$-vertex graph so we cannot
  % assign a probability to each edges existence, instead placing $m$
  % edges at random in the graph.}
  The system is then advanced step-by-step based on the following
  procedure:

@@

  \noindent where $d_i$ is the degree of vertex $v_i$, $E(G)$ is the
  set of edges in the graph, and $\kappa \in (0, \infty)$ is a model
  parameter affecting the influence degrees have on the probability
  distribution in Step \ref{step:prob}. 
%
  That is, taking $\lim\limits{\kappa \rightarrow 0}$, we recover
  ``pure'' preferential attachment, and probability scales directly
  with degree,
  %%% YGK immediately? after a short time ? at steady state ?  if at
  %%% steady state say it ?  ATH This refers to the dynamics at all
  %%% times. At every step, we need some probability of choosing a
  %%% given vertex v_i: if k -> 0, this probability is proportional to
  %%% deg(v_i), and if k -> infinity this probability is uniform over
  %%% all vertices (and degree has no effect. Valid at every step.
  while
  $\lim\limits_{\kappa \rightarrow \infty} P(v_k = v_l) = \frac{1}{n}
  \; \forall \; l$, and individual degrees have no effect. A single
  step of this evolution process is illustrated in
  Fig. ({\ref{fig:step-illustration}}). Note that this can also be
  represented as a graph in which only one edge is permitted between
  each pair of nodes, but with edge weights signifying the number, or
  strength, of connections between nodes.
  %%% YGK you mean individual initial degrees ?
%%%
  \par

@@

  \begin{figure}[th!]
    \vspace{-5mm} \centering
    \begin{subfigure}[t]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{lopsided-degs-a}
      \subcaption{\label{fig:lopsided-init}}
    \end{subfigure} %
    \begin{subfigure}[t]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{erdos-degs-a}
      \subcaption{\label{fig:erdos-init}}
    \end{subfigure}
    \caption{Evolution of the multigraph model's sorted degree
      sequence. Two distinct transients are shown, initialized with
      (a) an Erd\H{o}s-R\'{e}nyi random graph with $m = 5,050$ edges
      and (b) a graph in which half the vertices are isolated, and
      half uniformly share $m = 5,050$ edges. Both approach the same
      ultimate stationary sequence. To smoothen the evolution of this
      stochastic system, here we plot the instantaneous average of twenty
      simulations. \label{fig:dse}}
    % Multigraph evolution dynamics for an $n=100$ vertex graph
    % observed through the degree sequence (after sorting the nodes by
    % degree)
  \end{figure}

  \begin{figure}[th!]
    % \vspace{-5mm}
    % \centering
    % \begin{subfigure}[t]{0.49\textwidth}
    %   \centering
    %   \includegraphics[width=\textwidth]{triangle-slaving}
    %   \subcaption{\label{fig:sv-n2}}
    % \end{subfigure} %
    % \begin{subfigure}[t]{0.49\textwidth}
    \centering
    \includegraphics[width=0.8\textwidth]{triangle-slaving-n2-n3}
    % \subcaption{\label{fig:sv-n3}}
    % \end{subfigure}
    \caption{Evolution of higher-order network statistics (here, the
      total triangle count) for three different network
      initializations. They are quickly ($O(n^2)$ steps) drawn to a slow
      manifold on which they slowly evolve over ($O(n^3)$ steps) to a
      stationary state. \label{fig:sv}}
  \end{figure}
  %%% 
  %%% YGK maybe you can draw a short horizontal broken line at the "end"
  %%% of the graph, where "presumably" the steady state value is
  %%% extending from the boundary of the plot "inward" in the plot maybe
  %%% a thick black broken line ?

@@

  The lifting operator performs the inverse role of restriction:
  translating coarse variables into full networks.
  % 
  This is clearly a one-to-many mapping.
  % 
  Specifically in the context of our model, we map vectors of polynomial
  coefficients to full networks in a two-stage process.
  % 
  First, the coefficient vector is used to recreate a sorted degree
  sequence by evaluating the full polynomial at integer values and
  rounding to the nearest degree, as depicted in
  Fig. (\ref{fig:cpi-diagram}).
  % 
  If the sum of the resulting degree sequence is odd, a single degree is
  added to the largest value to ensure the sequence is graphical.
  % 
  Second, this degree sequence is used as input to any algorithm that
  produces a network consistent with a graphical degree sequence (here
  we typically used a Havel-Hakimi algorithm, creating a graph whose
  degree sequence matches that of the input
  \cite{havel_remark_1955,hakimi_realizability_1962}).
  % 
  While the canonical Havel-Hakimi method produces simple graphs, it is
  not difficult to extend it to multigraphs by allowing the algorithm to
  wrap around the sequence, producing multiple edges and self-loops.
  %%% YGK is this you, Alexander, is it Balazs, did you read it
  %%% somewhere, is there a reference ?  ATH This is me, but Balazs
  %%% confirmed this in his comments when he said that Havel Hakimi was
  %%% even overly-complex for this case as there are even simpler
  %%% methods one could use. I think it is easy to prove.
  The lifting procedure is represented by the red arrows of
  Fig. (\ref{fig:cpi-diagram}).

@@

  Note that Step (\ref{cpi:heal}) is intended to allow a singularly
  perturbed quantity to approach its slow manifold.
  % 
  Upon initializing a new full system (here, a new network) in Step
  (\ref{cpi:init}), higher order observables (for example, the triangle
  count) will be far from the values they would attain in a detailed
  simulation.
  % 
  If the coarse system model closes with our chosen variables, then
  either (a) these quantities do not affect the dynamics of the degree
  sequence; or (b) they do, but then by hypothesis these quantities
  quickly evolve to functions of the selected coarse variables (i.e.,
  they approach the slow manifold).
  % 
  In the second case, after a short interval of ``healing'' they will be
  drawn to the expected trajectory, after which we begin sampling.
  % 
  This is analogous to Fig. (\ref{fig:sv}).
  % 
  The computational gains arise from the projective step,
  (\ref{cpi:proj}).
  % 
  Here, we advance the system $t_p$ steps at the cost of one evaluation of forward Euler, instead of $t_p$ direct detailed steps. \\
  %%% YGK do you say what your step counts are in your actual
  %%% computation ?  ATH added specifics into the paragraph below

  Results of the application of this general framework with the specific
  lifting and restriction operators previously outlined are shown in
  Fig. (\ref{fig:cpi-results}). We used an $n=100$-vertex graph with
  $m=50000$ edges and parameter value $\kappa=1$. We ran the model for a
  total of $10 \cdot n^3$ steps, with $t_h = 10 \cdot n^2$, $t_c = n^3$
  and $t_p = 50 \cdot n^2$.
  % 
  We see good agreement between the CPI-accelerated and normal systems,
  while reducing the number of detailed steps by one third.
  % 
  It is important to mention that this method was applied not to a
  single system instantiation,
  % t
  but to an ensemble of fifty such realizations.
  %%% YGK say how many ?
  %%% 
  %%% ATH done (fifty)
  This ensured that when properties such as the fitted polynomial
  coefficients were averaged over the ensemble they evolved smoothly
  despite the stochastic nature of the system; in effect, we are
  computing the ``expected network evolution'' averaged over consistent
  realizations.

@@

  %%% YGK can you consistently make F and d bold, so that they are
  %%% vectors ?
  %%% 
  %%% ATH done

@@

  \subsection{PCA}
  \label{sec:pca}
  % TODO (referenced fig below)

@@

  Additionally, we were interested in embeddings of different model
  trajectories.
  % 
  This dataset was generated by sampling two different model simulations
  %%% YGK why even intervals ?  at even intervals ATH I don't remember a
  %%% particular reason
  as they evolved ($N=2000$ points were sampled from each trajectory).
  % 
  The parameters $n$, $m$ and $\kappa$ were held constant at $200$,
  $20100$ and $1.0$ respectively, but one graph was initialized as an
  Erd\H{o}s-R\'{e}nyi random graph (Fig. (\ref{fig:erdos-init})), while
  the other was initialized as a ``lopsided'' graph
  (Fig. (\ref{fig:lopsided-init})).
  % 
  Every $500$ steps the graph would be recorded till $N$ snapshots were
  taken of each trajectory, for a total of $1000000$ steps.
  % 
  Letting $G_e(t)$ refer to the Erd\H{o}s-R\'{e}nyi-initialized system
  at step $t$, and $G_l(t)$ the lopsided-initialized system, the
  embedding $\gamma$ was used to create points
  $\gamma(G_e(t)) = v_e(t) \in \mathbb{R}^8$ and
  similarly $\gamma(G_l(t)) = v_l(t) \in \mathbb{R}^8$. \\

  DMAPS was then applied to this set of $2N$ points, and the
  three-dimensional embedding using $\Phi_1$, $\Phi_2$ and $\Phi_3$ is
  shown in Fig. (\ref{fig:dmaps-results}).
  % 
  At first, the two trajectories are mapped to distant regions in
  $\mathbb{R}^3$ due to their different initial conditions.
  % 
  They evolve along different ``regions'' of the embedding, taking two
  distinct trajectories on their approach to the stationary state.
  % 
  % The particular eigenvector each trajectory travels along is also
  % significant: as $G_e(t)$ was mapped to $\Phi_1$ and $G_l(t)$ to
  % $\Phi_2$, we can infer that the dynamics of the
  % Erd\H{o}s-R\'{e}nyi-initialized system occur more slowly than those
  % of the lopsided-initialized system. Indeed, referring back to
  % Fig. (\ref{fig:dse}), we see that the lopsided network descends more
  % rapidly to the stationary state.
  Eventually, their embeddings meet as they arrive at this final, shared
  state, each asymptotically along opposite sides of the slowest
  eigenvector of the linearization of the steady state.
  % 
  DMAPS thus proves useful in elucidating both geometric and dynamic
  features of the system as shown in Figs. (\ref{fig:dmaps-rk}) and (\ref{fig:dmaps-results}). \\

@@

  The equation free framework was successfully applied to the
  edge-conservative preferential attachment model, accelerating
  simulations through CPI and locating stationary states through coarse
  Newton-GMRES. This indicates potential avenues for improving
  simulation times of other complex network models, such as those
  found in epidemiology.
  % 
  Additionally, an underlying two-dimensional description was
  uncovered via PCA and DMAPS. These automated methods of
  dimensionality-reduction are quite general, and can in principle be
  applied in a wide range of network settings to probe hidden
  low-dimensional structure. For an application in the setting of
  labeled nodes, see ({\cite{kattis_modeling_2015}}). \\

@@

  %% This poses several problems to the investigator. First, the
  %% interactions themselves may be complicated functions that are
  %% expensive to compute and dependent on several network
  %% parameters. This, coupled with the large network sizes inherent to
  %% many interesting problems makes numerical investigation
  %% computationally prohibitive.

  %% For instance, the simple model proposed in \ref{votingmodel}
  %% describing the adoption of opinions in a population is defined such
  %% that at each simulation step, a random individual is chosen and
  %% acted upon.  This precludes the use of traditional numerical
  %% techniques such as optimization and fixed-point location that could
  %% offer valuable insight into the network's behavior.  We expand

>>>>

\documentclass[epjST, final]{svjour} \usepackage{graphicx, subcaption,
  setspace, indentfirst, amsfonts, amsthm, empheq, color, soul}
\graphicspath{ {../figs/} }
\newcommand\ignore[1]{}

\title{Equation-free analysis of a dynamically evolving multigraph}
\author{Alexander Holiday \inst{1} \and Ioannis G. Kevrekidis
  \inst{1,2}} \institute{Department of Chemical and Biological
  Engineering, Princeton University, Princeton, New Jersey 08544, USA
  \and PACM, Princeton University, Princeton, New Jersey 08544, USA}

@@

  We study the edge-conservative preferential attachment model, a
  detailed mathematical analysis of which can be found in
  \cite{rath_time_2012} and \cite{rath_multigraph_2012}.
%
  The system evolves in discrete steps $t = 0,1,\ldots t_f$, and we
  denote the $n$-vertex graph at each point by $G_n(t)$.
%
  The initial state, $G_n(0)$, is an arbitrary distribution of $m$
  edges among the $n$ vertices; the total number of both edges and
  vertices is conserved.
%
  No restrictions are placed on this initial distribution: multiple
  edges and loops are permitted. 
  The system is then advanced step-by-step based on the following
  procedure:

@@

  \noindent where $d_i$ is the degree of vertex $v_i$, $E(G)$ is the
  set of edges in the graph, and $\kappa \in (0, \infty)$ is a model
  parameter affecting the influence degrees have on the probability
  distribution in Step \ref{step:prob}. 
%
  That is, taking $\lim\limits{\kappa \rightarrow 0}$, we recover
  ``pure'' preferential attachment, and probability scales directly
  with degree,
  while
  $\lim\limits_{\kappa \rightarrow \infty} P(v_k = v_l) = \frac{1}{n}
  \; \forall \; l$, and individual degrees have no effect. A single
  step of this evolution process is illustrated in
  Fig. ({\ref{fig:step-illustration}}). Note that this can also be
  represented as a graph in which only one edge is permitted between
  each pair of nodes, but with edge weights signifying the number, or
  strength, of connections between nodes.
  \par

@@

  \begin{figure}[th!]
    \vspace{-5mm} \centering
    \begin{subfigure}[t]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{lopsided-degs-a}
      \subcaption{\label{fig:lopsided-init}}
    \end{subfigure} %
    \begin{subfigure}[t]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{erdos-degs-a}
      \subcaption{\label{fig:erdos-init}}
    \end{subfigure}
    \caption{Evolution of the multigraph model's sorted degree
      sequence. Two distinct transients are shown, initialized with
      (a) an Erd\H{o}s-R\'{e}nyi random graph with $m = 5,050$ edges
      and (b) a graph in which half the vertices are isolated, and
      half uniformly share $m = 5,050$ edges. Both approach the same
      ultimate stationary sequence. To smoothen the evolution of this
      stochastic system, here we plot the instantaneous average of twenty
      simulations. \label{fig:dse}}
  \end{figure}

  \begin{figure}[th!]
    \centering
    \includegraphics[width=0.8\textwidth]{triangle-slaving-n2-n3}
    \caption{Evolution of higher-order network statistics (here, the
      total triangle count) for three different network
      initializations. They are quickly ($O(n^2)$ steps) drawn to a slow
      manifold on which they slowly evolve over ($O(n^3)$ steps) to a
      stationary state. \label{fig:sv}}
  \end{figure}

@@

  The lifting operator performs the inverse role of restriction:
  translating coarse variables into full networks.
  % 
  This is clearly a one-to-many mapping.
  % 
  Specifically in the context of our model, we map vectors of polynomial
  coefficients to full networks in a two-stage process.
  % 
  First, the coefficient vector is used to recreate a sorted degree
  sequence by evaluating the full polynomial at integer values and
  rounding to the nearest degree, as depicted in
  Fig. (\ref{fig:cpi-diagram}).
  % 
  If the sum of the resulting degree sequence is odd, a single degree is
  added to the largest value to ensure the sequence is graphical.
  % 
  Second, this degree sequence is used as input to any algorithm that
  produces a network consistent with a graphical degree sequence (here
  we typically used a Havel-Hakimi algorithm, creating a graph whose
  degree sequence matches that of the input
  \cite{havel_remark_1955,hakimi_realizability_1962}).
  % 
  While the canonical Havel-Hakimi method produces simple graphs, it is
  not difficult to extend it to multigraphs by allowing the algorithm to
  wrap around the sequence, producing multiple edges and self-loops.
  The lifting procedure is represented by the red arrows of
  Fig. (\ref{fig:cpi-diagram}).

@@

  Note that Step (\ref{cpi:heal}) is intended to allow a singularly
  perturbed quantity to approach its slow manifold.
  % 
  Upon initializing a new full system (here, a new network) in Step
  (\ref{cpi:init}), higher order observables (for example, the triangle
  count) will be far from the values they would attain in a detailed
  simulation.
  % 
  If the coarse system model closes with our chosen variables, then
  either (a) these quantities do not affect the dynamics of the degree
  sequence; or (b) they do, but then by hypothesis these quantities
  quickly evolve to functions of the selected coarse variables (i.e.,
  they approach the slow manifold).
  % 
  In the second case, after a short interval of ``healing'' they will be
  drawn to the expected trajectory, after which we begin sampling.
  % 
  This is analogous to Fig. (\ref{fig:sv}).
  % 
  The computational gains arise from the projective step,
  (\ref{cpi:proj}).
  % 
  Here, we advance the system $t_p$ steps at the cost of one evaluation of forward Euler, instead of $t_p$ direct detailed steps. \\

  Results of the application of this general framework with the specific
  lifting and restriction operators previously outlined are shown in
  Fig. (\ref{fig:cpi-results}). We used an $n=100$-vertex graph with
  $m=50000$ edges and parameter value $\kappa=1$. We ran the model for a
  total of $10 \cdot n^3$ steps, with $t_h = 10 \cdot n^2$, $t_c = n^3$
  and $t_p = 50 \cdot n^2$.
  % 
  We see good agreement between the CPI-accelerated and normal systems,
  while reducing the number of detailed steps by one third.
  % 
  It is important to mention that this method was applied not to a
  single system instantiation,
  but to an ensemble of fifty such realizations.
  This ensured that when properties such as the fitted polynomial
  coefficients were averaged over the ensemble they evolved smoothly
  despite the stochastic nature of the system; in effect, we are
  computing the ``expected network evolution'' averaged over consistent
  realizations.

@@

@@

  \subsection{PCA}
  \label{sec:pca}

@@

  Additionally, we were interested in embeddings of different model
  trajectories.
  % 
  This dataset was generated by sampling two different model simulations
  as they evolved ($N=2000$ points were sampled from each trajectory).
  % 
  The parameters $n$, $m$ and $\kappa$ were held constant at $200$,
  $20100$ and $1.0$ respectively, but one graph was initialized as an
  Erd\H{o}s-R\'{e}nyi random graph (Fig. (\ref{fig:erdos-init})), while
  the other was initialized as a ``lopsided'' graph
  (Fig. (\ref{fig:lopsided-init})).
  % 
  Every $500$ steps the graph would be recorded till $N$ snapshots were
  taken of each trajectory, for a total of $1000000$ steps.
  % 
  Letting $G_e(t)$ refer to the Erd\H{o}s-R\'{e}nyi-initialized system
  at step $t$, and $G_l(t)$ the lopsided-initialized system, the
  embedding $\gamma$ was used to create points
  $\gamma(G_e(t)) = v_e(t) \in \mathbb{R}^8$ and
  similarly $\gamma(G_l(t)) = v_l(t) \in \mathbb{R}^8$. \\

  DMAPS was then applied to this set of $2N$ points, and the
  three-dimensional embedding using $\Phi_1$, $\Phi_2$ and $\Phi_3$ is
  shown in Fig. (\ref{fig:dmaps-results}).
  % 
  At first, the two trajectories are mapped to distant regions in
  $\mathbb{R}^3$ due to their different initial conditions.
  % 
  They evolve along different ``regions'' of the embedding, taking two
  distinct trajectories on their approach to the stationary state.
  % 
  Eventually, their embeddings meet as they arrive at this final, shared
  state, each asymptotically along opposite sides of the slowest
  eigenvector of the linearization of the steady state.
  % 
  DMAPS thus proves useful in elucidating both geometric and dynamic
  features of the system as shown in Figs. (\ref{fig:dmaps-rk}) and (\ref{fig:dmaps-results}). \\

@@

  The equation free framework was successfully applied to the
  edge-conservative preferential attachment model, accelerating
  simulations through CPI and locating stationary states through coarse
  Newton-GMRES. This indicates potential avenues for improving
  simulation times of other complex network models, such as those
  found in epidemiology.
  % 
  Additionally, an underlying two-dimensional description was
  uncovered via PCA and DMAPS. These automated methods of
  dimensionality-reduction are quite general, and can in principle be
  applied in a wide range of network settings to probe hidden
  low-dimensional structure. For an application in the setting of
  labeled nodes, see ({\cite{kattis_modeling_2016}}). \\

@@
